---
title: "SYS 6018: Data Mining Final Project" 
author: "Lea Jih-Vieira, Mitch Whalen"
format: sys6018hw-html
---

```{r config, include=FALSE}
# Set global configurations and settings here
knitr::opts_chunk$set()                 # set global chunk options
ggplot2::theme_set(ggplot2::theme_bw()) # set ggplot2 theme
```

# Required R packages {.unnumbered .unlisted}

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse) # functions for data manipulation  
# add additional libraries here
```

# Introduction

Describe background and goal of analysis...

- why we are doing it
- scenario (foraging for mushrooms, what is the probability the mushroom you picked up is poisonous)

# Load Data

```{r}
df = read_csv("mushroom_cleaned.csv", col_names = TRUE)
```
```{r}
head(df)
```

# Data Description

List columns here and their descriptions (sourced from UCI website)...

# Data Cleaning

Explanation here...

I think the Kaggle page already went through some data cleaning so we can talk about that here and then see if we need to do any additional cleaning in addition to that.

```{r}
# data cleaning code here
```

# Model Development

Brief description of models chosen and why, they will be combined into ensemble and see how that compares...

## Model 1: Logistic Regression

```{r}
# model 1 code here
```

## Model 2: Tree-Based Method

```{r}
# model 2 code here
```

## Model 3: Support Vector Machine
- logistic and trees produce probabilities, SVM doesn't
- can make SVM into a probability using plat score (??) or something
- just need to be careful if ensembling because not all the same scale

```{r}
# model 3 code here
```

## Model 4: Ensemble

- can also discuss maybe only need one model versus ensemble

```{r}
# model 4 code here
```

# Model Evaluation

Evaluate models here...

misclassification analysis

# Discussion

Discuss results here (maybe can be combined with Model Evaluation)...

# Conclusion

Conclusions here...

# Limitations

Maybe talk about limitations?? here...