---
title: "SYS 6018: Data Mining Final Project" 
author: "Lea Jih-Vieira, Mitch Whalen"
format: sys6018hw-html
---

```{r config, include=FALSE}
# Set global configurations and settings here
knitr::opts_chunk$set()                 # set global chunk options
ggplot2::theme_set(ggplot2::theme_bw()) # set ggplot2 theme
```

# Required R packages {.unnumbered .unlisted}

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse) # functions for data manipulation  
# add additional libraries here
```

# Introduction

Describe background and goal of analysis...

# Load Data

```{r}
#df = read_csv("mushroom_cleaned.csv", col_names = TRUE)
```
```{r}
#head(df)
```

# Data Description

List columns here and their descriptions (sourced from UCI website)...

# Data Cleaning

Explanation here...

```{r}
# data cleaning code here
```

# Model Development

Brief description of models chosen and why, they will be combined into ensemble and see how that compares...

## Model 1: Logistic Regression

```{r}
# model 1 code here
```

## Model 2: Tree-Based Method

```{r}
# model 2 code here
```

## Model 3: Support Vector Machine

```{r}
# model 3 code here
```

## Model 4: Ensemble

```{r}
# model 4 code here
```

# Model Evaluation

Evaluate models here...

# Discussion

Discuss results here (maybe can be combined with Model Evaluation)...

# Conclusion

Conclusions here...

# Limitations

Maybe talk about limitations?? here...