---
title: "SYS 6018: Data Mining Final Project" 
author: "Lea Jih-Vieira, Mitch Whalen"
format: sys6018hw-html
---

```{r config, include=FALSE}
# Set global configurations and settings here
knitr::opts_chunk$set()                 # set global chunk options
ggplot2::theme_set(ggplot2::theme_bw()) # set ggplot2 theme
```

# Required R packages {.unnumbered .unlisted}

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse) # functions for data manipulation  
# add additional libraries here
```

# Introduction

Many people enjoy mushroom foraging as a hobby and a lifestyle, hunting through woods to find ingredients for their next meal. However, foraging for mushrooms is not a casual activity by any means. Foraging requires vast knowledge of the wild mushrooms that grow in a given area and how to identify them ([Incredible Mushrooms](https://www.incrediblemushrooms.com/wild-mushroom-foraging.html)). It is imperative that a forager can tell the difference between an edible mushroom and a poisonous one, especially when the poisonous varieties can look almost identical to their edible counterparts ([Incredible Mushrooms](https://www.incrediblemushrooms.com/wild-mushroom-foraging.html)).

This report aims to provide foragers a model-based method of determining the probability that a foraged mushroom is poisonous or not. To accomplish this, we will utilize a dataset of macromorphological mushroom characteristics that a forager could observe without the need for more advanced technologies. By providing a model-based method to determine the probability of mushroom toxicity using easily observable features, we hope to provide foragers a more reliable and accurate method to identify mushrooms.

# Load Data

```{r}
df = read_csv("mushroom_cleaned.csv", col_names = TRUE)
```
```{r}
head(df)
```

# Data Description

```{r}
# View the structure of the data
str(df)
```
Our data set has 54,035 instances of 9 variables (Cap diameter, cap shape, gill attachment, gill color, stem height, stem width, stem color, season, and class). Class is our response variable where 0 refers to edible and 1 refers to poisonous.

```{r}
# Summary statistics
summary(df)
```
Based on the summary statistics of our data, we can see that the class variable has a mean of 0.5492, thus the data set is relatively balanced. Cap diameter, stem height, and stem width are quantitative predictor variables. Whereas cap shape, gill attachment, gill color, stem color, and season are categorical predictor variables which will require treatment in data cleaning. 

# Data Cleaning

Explanation here...

I think the Kaggle page already went through some data cleaning so we can talk about that here and then see if we need to do any additional cleaning in addition to that.

```{r}
# data cleaning code here
```

# Model Development

Brief description of models chosen and why, they will be combined into ensemble and see how that compares...

## Model 1: Logistic Regression

```{r}
# model 1 code here
```

## Model 2: Tree-Based Method

```{r}
# model 2 code here
```

## Model 3: Support Vector Machine
- logistic and trees produce probabilities, SVM doesn't
- can make SVM into a probability using plat score (??) or something
- just need to be careful if ensembling because not all the same scale

```{r}
# model 3 code here
```

## Model 4: Ensemble

- can also discuss maybe only need one model versus ensemble

```{r}
# model 4 code here
```

# Model Evaluation

Evaluate models here...

misclassification analysis

# Discussion

Discuss results here (maybe can be combined with Model Evaluation)...

# Conclusion

Conclusions here...

# Limitations

Maybe talk about limitations?? here...